\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-01 doc/publications/plos16/plos16.tex
a few changes

\section{Introduction}
\marginpar{
we have parallelize and parallelise}

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
\marginpar{why not just error rate}
and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
these parameters can have on the prediction accuracy of the model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\marginpar{Rob can you condense this down into what is relevant for this paper and not explained elsewhere in the literature}
\subsection{Overview of Random Forests}

@@

The average prediction error for an individual tree $h(X; B)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage disequilibrium?}

@@

\subsubsection{Synthetic data} 
\marginpar{these are not actually encoded as factors}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
  have a very poor
\marginpar{is this true}
  fit, and be useless for prediction, and still have a useful ranking of variable importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{replicated in methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.
\marginpar{in the terminology of statistics, these are all main effects. There are no interaction terms}

@@

\marginpar{the accuracy hardly changes}
Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

\marginpar{wrong figure cited}
Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.

@@

\marginpar{figure needs bigger text}
\begin{figure}[tbhp] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
    \caption{Variable importance of top 20 variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oob.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
    forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model. The dotted line represents the original linear
  model coefficients.
 {\bf b)} OOB (out of bag) classification error
  estimate as function of \mtry\ fraction.}
  \label{figure:synth}
\end{figure}

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\

@@

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\

@@

\marginpar{can we remove this table and just rely on the figure?}
\begin{table}[!ht]
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees with specified \mtry\ fraction using given number of CPU cores.}}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
\cline{3-6}
\multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
\hline
\multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
                                        & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
                                        & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
                                        & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{table10}
\end{table}

@@

>>>>

\section{Introduction}
\marginpar{
We have parallelize and parallelise.
US spelling. Rob
}

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2.
Aidan} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
these parameters can have on the prediction accuracy of the model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview of Random Forests}

@@

The average prediction error for an individual tree $h(X; B)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss, Rob}
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
conditions caused by the interaction of large groups of SNPs. 
\marginpar{check this. Denis}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage disequilibrium? Denis}

@@

\subsubsection{Synthetic data} 
\marginpar{check details of simulation. Piotr}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal variables with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trial with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R terms from the {\sc randomForest} package
\cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate the
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
\cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a random forest model to
  have a very poor
\marginpar{is this true. If unsupported, remove it. Rob}
  fit, and be useless for prediction, and still have a useful ranking of variable importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in {\sc randomForest} by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets \texttt{maxDepth}
    to 4 in their classification example. This is wrong. The
    individual trees should be grown to a large size to ensure that
    they give an unbiased estimate.}.
  In the TCGA example the RandomForests code produced trees with depths of 30+. 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{replicated in methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple additive features.

@@

Fig.~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

\marginpar{wrong figures cited. Rob}
Fig.~\ref{figure:synth}b illustrates that increasing \mtry\ also improves the OOB (out of bag) classification error.
%Piotr can you please explain the remaining error on the synthetic data
By $mtry=0.8$ the error is converging to X, which is due to Y.
\marginpar{what are ``X'' and ``Y''. Piotr}

@@

\marginpar{figure needs bigger text}
\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oder.png} 
    \caption{Variable importance of top 20 variables.} 
    \label{fig7:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=6cm]{./figs/var_oob.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
\caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
    forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  {\bf a)} Variable
  importance of top 20 variables normalized to the sum of the linear model coefficients. {\bf Others} represents
  aggregated importance of variables not present in the linear model. The dotted line represents the original linear
  model coefficients.
 {\bf b)} OOB (out of bag) classification error
  estimate as function of \mtry\ fraction.}
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error  & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077  &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.06       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.06       & 7 mins           & --nthreads 20                                                     \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063  & 1hr 58min        & 4GB \\

@@

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046  & 4hr 0min                                             & 4GB \\

@@

@@

% \marginpar{can we remove this table and just rely on the figure?}
% \begin{table}[!ht]
% \centering
% \caption{
% {\bf Time in seconds to build a random forest of 100
%   trees with specified \mtry\ fraction using given number of CPU cores.}}
% \begin{tabular}{|l|l|l|l|l|l|}
% \hline
% \multicolumn{2}{|l|}{\multirow{2}{*}{}}             & \multicolumn{4}{c|}{Number of CPU cores} \\
% \cline{3-6}
% \multicolumn{2}{|l|}{}                              & \bf{16} & \bf{32} & \bf{64} & \bf{128} \\
% \hline
% \multirow{4}{*}{\mtry\ (fraction)}        & \bf{0.10} & 708.3   & 340.0   & 186.0   & 134.5 \\
%                                         & \bf{0.25} & 1134.9  & 535.0   & 270.8   & 179.6 \\
%                                         & \bf{0.50} & 1724.1  & 856.6   & 428.7   & 243.8 \\
%                                         & \bf{0.80} & 2313.5  & 1125.7  & 569.1   & 364.6 \\
% \hline
% \end{tabular}
% \begin{flushleft} 
% \end{flushleft}
% \label{table10}
% \end{table}
